---
title: "Function W3 PCA"
author: "Sean Sutton"
date: '2022-10-26'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(ggplot2)
library(MASS)
library(tidyverse)
library(mvtnorm)
```

## Demo data
```{r}
aircraft_raw <- read.csv("dat/aircraft.csv")

air <- aircraft_raw %>%
  dplyr::select(3:8) %>%
  mutate_all(log10)

head(air)
```

## Eigenvectors and values
```{r}
pc <- prcomp(air)

## Eigenvectors
pc$rotation
```
```{r}
## Eigenvalues
pc$sdev ^ 2
```
### Alternative method via covariance matrix
```{r}
cov_mat <- cov(air)
eig <- eigen(cov_mat)
eig$values
eig$vectors
```
## Outputs of pr_comp
pc$sdev - the standard deviation accounted for by each PC score
Note this adds up to the total variance of each individual component
```{r}
sum(pc$sdev ** 2)
sum(diag(cov(air)))
```
pc$rotation - the PC loadings for each variable
pc$center - the mean value of each column in the initial data
pc$scale - whether the data has been scaled
pc$x - the PC scores for each observation

## Scree plot - Cumulative contribution to total variance
```{r}
scree_cum_prop_variance <- function(pc_obj, title="") {
  cum_contri <- cumsum(pc_obj$sdev ** 2)
  plot_df <- data.frame(idx = 1:6, cum_contri = cum_contri)
  ggplot(data = plot_df, aes(idx, cum_contri)) +
    geom_line() +
    xlab("PC Score") +
    ylab("Cumulative contribution to total variance") +
    ggtitle(title)
}
```
e.g.
```{r}
scree_cum_prop_variance(prcomp(air), "Unscaled Logged PC Scores")
```
## Checking the distribution of a PC score (smoother histogram (density) of the PC scores)
```{r}
ggplot(data.frame(pc$x), aes(PC1)) +
  geom_density()
```
## Colourcoded scatteplot of pc scores
```{r}
ggplot(data.frame(pc$x), aes(PC1, PC2, color=as.factor(aircraft$Period))) +
  geom_point() +
  geom_density_2d(color=1, alpha=0.5) + ## Density
  labs(color = "Period") ## Label for type of categories
```


## Biplot
```{r}
biplot(pc, col=c("white", "blue"))
```

## Generating samples from the normal distribution
```{r}
Sigma0 = matrix(c(
 3.1386518, 0.38872659, 0.6178228, 1.7576847, 0.77433973, 0.7508789,
 0.3887266, 1.21417949, 0.1941453, 0.4518920, 0.01236855, 0.2155466,
 0.6178228, 0.19414529, 1.2437919, 0.5970320, 0.15151088, 0.2665018,
 1.7576847, 0.45189196, 0.5970320, 1.7083497, 0.52685668, 0.7109476,
 0.7743397, 0.01236855, 0.1515109, 0.5268567, 0.53406192, 0.2299193,
 0.7508789, 0.21554658, 0.2665018, 0.7109476, 0.22991933, 0.6642375),
 byrow = TRUE, nrow = 6 )
```

```{r}
n = 250 # number of observations
Sigma = Sigma0 # covariance matrix
mu = rep(0, 6) # the means of each variable
normal_sample <- mvrnorm(n=n, mu = mu, Sigma = Sigma)
summary(normal_sample)
```

## Generating samples from the t-distribution
```{r}
Sigma0sym = 0.5* (Sigma0 + t(Sigma0))
Xmat = rmvt(100, sigma = Sigma0sym, df = 6)
summary(Xmat)
```

## Plot density by group
```{r}
multiple_density_plot = function(data, col, group){
  # usage:
  # data: dataframe
  # col: column to ploy (name of column)
  # group: the variable to group by (name of column)
  return(
    ggplot(data, aes(data[,col], group = group) ) +
    geom_density( aes(fill = group), alpha = 0.4 ) +
    theme( legend.position = "bottom") +
    labs(x = deparse(substitute(col)), y = "Density", fill = deparse(substitute(group)))
  )
}

multiple_density_plot(data.frame(pc$x), 1, as.factor(aircraft$Period))
```

## Plot density iteratively
```{r}

```