---
title: "Function W3 PCA"
author: "Sean Sutton"
date: '2022-10-26'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(ggplot2)
library(MASS)
library(tidyverse)
library(mvtnorm)
```

## Demo data
```{r}
aircraft_raw <- read.csv("dat/aircraft.csv")

air <- aircraft_raw %>%
  dplyr::select(3:8) %>%
  mutate_all(log10)

head(air)
```

## Eigenvectors and values
```{r}
pc <- prcomp(air)

## Eigenvectors
pc$rotation
```
```{r}
## Eigenvalues
pc$sdev ^ 2
```
### Alternative method via covariance matrix
```{r}
cov_mat <- cov(air)
eig <- eigen(cov_mat)
eig$values
eig$vectors
```
## Outputs of pr_comp
pc$sdev - the standard deviation accounted for by each PC score
Note this adds up to the total variance of each individual component
```{r}
sum(pc$sdev ** 2)
sum(diag(cov(air)))
```
pc$rotation - the PC loadings for each variable
pc$center - the mean value of each column in the initial data
pc$scale - whether the data has been scaled
pc$x - the PC scores for each observation

## Scree plot - Cumulative contribution to total variance
```{r}
scree_cum_prop_variance <- function(pc_obj, title="") {
  cum_contri <- cumsum(pc_obj$sdev ** 2)
  plot_df <- data.frame(idx = 1:6, cum_contri = cum_contri)
  ggplot(data = plot_df, aes(idx, cum_contri)) +
    geom_line() +
    xlab("PC Score") +
    ylab("Cumulative contribution to total variance") +
    ggtitle(title)
}
```
e.g.
```{r}
scree_cum_prop_variance(prcomp(air), "Unscaled Logged PC Scores")
```
## Checking the distribution of a PC score (smoother histogram (density) of the PC scores)
```{r}
ggplot(data.frame(pc$x), aes(PC1)) +
  geom_density()
```
## Colourcoded scatteplot of pc scores
```{r}
ggplot(data.frame(pc$x), aes(PC1, PC2, color=as.factor(aircraft$Period))) +
  geom_point() +
  geom_density_2d(color=1, alpha=0.5) + ## Density
  labs(color = "Period") ## Label for type of categories
```


## Biplot
```{r}
biplot(pc, col=c("white", "blue"))
```